---
title: "Coursera - Reproducible Research: Peer Assessment 1"
author: "Hans W Hiser"
date: "`r Sys.Date()`"
output: html_document
---
## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. [1]
## Data
The data was downloaded from the course website and was stored locally.
```{r}
library(knitr)
library(ggplot2)
library(data.table)
opts_chunk$set(echo = TRUE, results = 'hold')
```
The file was loaded and below you can see a print out of a few columns of the data.
```{r}
#activity <- unzip("activity.zip")
act_data <- read.csv("E://RDir/RRPAssessment1/activity.csv", header=TRUE, sep=",")
str(act_data)
```
## Peer Graded Assignment Course Project 1
### Process the data
Convert some of the vectors to appropraite forms.

```{r}
act_data$date <- as.Date(act_data$date, format="%Y-%m-%d")
act_data$interval <- as.factor(act_data$interval)
```

Post converting the column classes print the structure of the data

```{r}
str(act_data)
```
Print the header of the dataset

```{r}
head(act_data, 10)
```
### What is the mean and median of the total number of steps per day?
Calculate the total number of steps taken in a day.
```{r}
steps_per_day <- aggregate(steps ~ date, data=act_data, FUN=sum)
colnames(steps_per_day) <- c("date", "steps")
```

Print the header of the data frame with steps aggregated per day.

```{r}
head(steps_per_day, 10)
```

### Make a histogram of the totla number of steps taken in a day.

```{r}
ggplot(steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 1000) + 
  labs(title = "Histogram - Steps Taken Per Day", x = "Steps Per Day", y = "Frequency")
```

Calculate and report the mean and median of steps taken per day.

```{r}
mean_steps_per_day <- mean(steps_per_day$steps)
mean_steps_per_day
median_steps_per_day <- median(steps_per_day$steps)
median_steps_per_day
```

### What is the average daily activity pattern?
Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
steps_per_interval <- aggregate(steps ~ interval, data = act_data, FUN = mean, na.rm = TRUE)
steps_per_interval$interval <- as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```

Print the header of the data frame with steps aggregated per interval

```{r}
head(steps_per_interval, 10)
```

```{r}
ggplot(steps_per_interval, aes(x = interval, y = steps)) + 
  geom_line(col = "blue", linewidth = 1) + 
  labs(title = "Average Daily Activity Pattern", x = "Interval", y = "Steps")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_interval <- steps_per_interval[which.max(steps_per_interval$steps),]
max_interval
```

Imputing missing values

```{r}
missing_values <- sum(is.na(act_data$steps))
missing_values
```

### Devise a strategy for filling in all of the missing values in the data set. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

To populate missing values, we choose to replace them with the mean value at the same interval across days.

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
new_act_data <- act_data
index_of_na <- which(is.na(new_act_data$steps))
for (i in index_of_na) {
  new_act_data$steps[i] <- with(steps_per_interval, steps[interval = new_act_data$interval[i]])
}
```

Print the top rows of newly created dataset

```{r}
head(new_act_data, 10)
```

Given we have handled the missing values, let us check if the above strategy really worked out.

```{r}
new_missing_values <- sum(is.na(new_act_data$steps))
new_missing_values
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
new_steps_per_day <- aggregate(steps ~ date, data = new_act_data, FUN=sum)
colnames(new_steps_per_day) <- c("date", "steps")
ggplot(new_steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 1000) + 
  labs(title = "Histogram - Steps Taken Per Day", x = "Steps Per Day", y = "Frequency")
```

In order to find the impact of imputing the missing values, let us compute the mean and median of steps taken per day

```{r}
new_mean_steps_per_day <- mean(new_steps_per_day$steps)
new_mean_steps_per_day
new_median_steps_per_day <- median(new_steps_per_day$steps)
new_median_steps_per_day
```

As we can see both mean and median of steps taken per day became exactly same which was not the case prior to imputing missing values

### Are there differences in activity patterns between weekdays and weekends?
#### Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Let us first add a factor variable to identify the given date as Weekday or Weekend
```{r}
dt <- data.table(new_act_data)
dt[, weekday := ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")]
dt$weekday <- as.factor(dt$weekday)
dt$interval <- as.integer(levels(dt$interval)[dt$interval])
head(dt, 10)
```

### Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
steps_per_weekday <- aggregate(steps ~ interval+weekday, data = dt, FUN = mean)
ggplot(steps_per_weekday, aes(x = interval, y = steps)) + 
  geom_line(col = "blue", linewidth = 1) + 
  facet_wrap(~ weekday, nrow=2, ncol=1) + 
  labs(x = "Interval", y = "Number of Steps")
```

Looking at the above graph we notice that the activity on weekdays has the highest peak (> 300) compared to all intervals and only one other peak that touches 100. On the contrary, weekend intervals have more peaks over a hundred than weekday. May be the person from whomever the data is collected is engaged in more active life style during weekends compared to weekdays.